# Community composition

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE}
load("data/data.Rdata")
genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))

remove_sled <- read_csv("data/sample_metadata.csv") %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  mutate(country = if_else(country == "Shangai domestic dog", "Shangai", country)) %>% 
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_type = case_when(
      breed == "Wolf"        ~ "wolf",
      country == "Greenland" ~ "sled_dog",
      TRUE                   ~ "domestic"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```

## Taxonomy overview 

### Stacked barplot
```{r taxonomy_barplot_canids, fig.height=8, fig.width=20, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
    geom_bar(stat="identity", colour="white", linewidth=0.1) + 
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ country, scale="free", space="free")+
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```
```{r taxonomy_barplot, fig.height=10, fig.width=15, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(., genome_metadata, by = join_by(genome == genome)) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_grid(~factor(dog_type, labels=c("domestic" = "Domestic dogs", "sled_dog" = "Sled dogs", "wolf" = "Wolves")), scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Number of Archaea phyla**

```{r arch, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull()
```
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage_nosp=100*count_nospecies/count_total) %>% 
  arrange(percentage_nosp) %>% 
  tt()
```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nmags <- nrow(genome_metadata)
nonspecies <- genome_metadata %>%
    filter(species == "") %>%
    nrow()
perct <- nonspecies*100/nmags
perct
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "") %>%
    nrow()
cat(nongenera)
```



### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,dog_type) %>%
  summarise(relabun=sum(count))
```

#### Origin: Dogs vs Sled dogs vs Wolves
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    Dog_mean = mean(relabun[dog_type == "domestic"] * 100, na.rm = T),
    Dog_sd = sd(relabun[dog_type == "domestic"] * 100, na.rm = T),
    Sled_dog_mean = mean(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Sled_dog_sd = sd(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Wolf_mean = mean(relabun[dog_type == "wolf"] * 100, na.rm = T),
    Wolf_sd = sd(relabun[dog_type == "wolf"] * 100, na.rm = T)
  ) %>%
  mutate(
    total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Domestic_dogs = str_c(round(Dog_mean, 3), "±", round(Dog_sd, 3)),
    Sled_dogs = str_c(round(Sled_dog_mean, 3), "±", round(Sled_dog_sd, 3)),
    Wolves = str_c(round(Wolf_mean, 3), "±", round(Wolf_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, total, Domestic_dogs, Sled_dogs, Wolves) %>%
  tt()
```

**Using CLR transformation**

```{r}
set.seed(0206)
phylo_counts <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_geno_meta <- genome_metadata %>%
  column_to_rownames(var="genome")%>%
  as.matrix()%>%
  tax_table()

physeq_genome_filt <- phyloseq(phylo_counts,phylo_geno_meta)
physeq_phylum <- tax_glom(physeq_genome_filt, taxrank = "phylum")
physeq_phylum_clr <- microbiome::transform(physeq_phylum, "clr")

df <- psmelt(physeq_phylum_clr)

results <- df %>%
  left_join(., sample_metadata, by=join_by("Sample"=="sample")) %>% 
  group_by(phylum) %>%
  do({
    model <- lm(Abundance ~ dog_type, data = .)
    as.data.frame(pairs(emmeans(model, ~ dog_type), adjust = "BH"))
  }) %>%
  ungroup()%>% 
  filter(p.value < 0.05) 

```
```{r}
results %>% 
  filter(contrast=="domestic - sled_dog")
```

```{r}
results %>% 
  filter(contrast=="domestic - wolf")
```

```{r}
results %>% 
  filter(contrast=="sled_dog - wolf")
```