# Community composition

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE, eval=FALSE}
load("data/data.Rdata")
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ country, scale="free", space = "free")+
 #   facet_nested(. ~ region+diet,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Bacteria phyla in wild individuals**

```{r phyla_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}

wild_samples <- sample_metadata %>% 
  filter(region=="Gipuzkoa") %>% 
  dplyr::select(sample) %>% 
  pull()

wild_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(wild_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% wild_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla captive animals**

```{r phyla_cap, comment="", echo=FALSE, message=FALSE, warning=FALSE}

captive_samples <- sample_metadata %>% 
  filter(region=="Nafarroa") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla before grass is included in the diet**

```{r phyla_pre, comment="", echo=FALSE, message=FALSE, warning=FALSE}

captive_pre_samples <- sample_metadata %>% 
  filter(diet=="Pre_grass") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_pre_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_pre_samples)) %>%
  as.data.frame() %>% 
  mutate(row_sum = rowSums(.)) %>%
  filter(row_sum != 0) %>%
  select(-row_sum)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_pre_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla after grass is included in the diet**

```{r phyla_post, comment="", echo=FALSE, message=FALSE, warning=FALSE}
captive_post_samples <- sample_metadata %>% 
  filter(diet=="Post_grass") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_post_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_post_samples)) %>%
  as.data.frame() %>% 
  mutate(row_sum = rowSums(.)) %>%
  filter(row_sum != 0) %>%
  select(-row_sum)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_post_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Number of Archaea phyla**

```{r arch, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Archaea phyla in wild individuals**

```{r arch_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% wild_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Archaea phyla before grass is included in the diet**

```{r arch_pre, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% captive_pre_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull()
```

**Archaea phyla after grass is included in the diet**

```{r arch_post, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% captive_post_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "s__") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull()

```
```{r nonspe_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "s__") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage=100*count_nospecies/count_total) %>% 
  tt()

```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nmags <- nrow(genome_counts)
nonspecies <- genome_metadata %>%
    filter(species == "s__") %>%
    nrow()
perct <- nonspecies*100/nmags
perct
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "g__") %>%
    nrow()
cat(nongenera)
```


### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,region, diet) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_boxplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    filter(phylum %in% phylum_arrange) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        theme(legend.position="none") +
        labs(y="Phylum",x="Relative abundance")
```

#### Origin: Wild vs Captive
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
    group_by(phylum) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Wild_mean=mean(relabun[diet=="Wild"]*100, na.rm=T),
              Wild_sd=sd(relabun[diet=="Wild"]*100, na.rm=T),
              Captive_mean=mean(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Captive_sd=sd(relabun[diet=="Pre_grass"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,3),"±",round(total_sd,3)),
           Wild=str_c(round(Wild_mean,3),"±",round(Wild_sd,3)),
           Captive=str_c(round(Captive_mean,3),"±",round(Captive_sd,3))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(phylum,total,Wild,Captive)
```

#### Origin and diet
```{r taxonomy_phylum_summary_origin_diet, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
    group_by(phylum) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Wild_mean=mean(relabun[diet=="Wild"]*100, na.rm=T),
              Wild_sd=sd(relabun[diet=="Wild"]*100, na.rm=T),
              Pre_grass_mean=mean(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Pre_grass_sd=sd(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Post_grass_mean=mean(relabun[diet=="Post_grass"]*100, na.rm=T),
              Post_grass_sd=sd(relabun[diet=="Post_grass"]*100, na.rm=T))  %>%
    mutate(total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Wild=str_c(round(Wild_mean,2),"±",round(Wild_sd,2)),
           Pre_grass=str_c(round(Pre_grass_mean,6),"±",round(Pre_grass_sd,6)),
           Post_grass=str_c(round(Post_grass_mean,2),"±",round(Post_grass_sd,2))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(phylum,total,Wild,Pre_grass,Post_grass)
```



```{r taxonomy_jitterplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    left_join(genome_metadata %>% select(phylum,phylum) %>% unique(),by=join_by(phylum==phylum)) %>%
#    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    filter(phylum %in% phylum_arrange[1:20]) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~diet)+
        theme_minimal() + 
        labs(y="phylum", x="Relative abundance", color="Phylum")
```

## Taxonomy boxplot

### Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family, diet,region) %>%
  summarise(relabun=sum(count))
family_summary$diet <- factor(family_summary$diet, levels=c("Pre_grass", "Post_grass", "Wild"))
```

#### Origin: Wild vs Captive
```{r taxonomy_family_summary_origin, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Wild_mean=mean(relabun[diet=="Wild"]*100, na.rm=T),
              Wild_sd=sd(relabun[diet=="Wild"]*100, na.rm=T),
              Cap_mean=mean(relabun[region=="Nafarroa"]*100, na.rm=T),
              Cap_sd=sd(relabun[region=="Nafarroa"]*100, na.rm=T))  %>%
    mutate(Total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Wild=str_c(round(Wild_mean,2),"±",round(Wild_sd,2)),
           Captive=str_c(round(Cap_mean,2),"±",round(Cap_sd,2))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(family,Total,Wild,Captive) %>% 
    paged_table()
```

#### Origin and Diet
```{r taxonomy_family_summary_type, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Wild_mean=mean(relabun[diet=="Wild"]*100, na.rm=T),
              Wild_sd=sd(relabun[diet=="Wild"]*100, na.rm=T),
              Pre_grass_mean=mean(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Pre_grass_sd=sd(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Post_grass_mean=mean(relabun[diet=="Post_grass"]*100, na.rm=T),
              Post_grass_sd=sd(relabun[diet=="Post_grass"]*100, na.rm=T))  %>%
    mutate(Total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Wild=str_c(round(Wild_mean,2),"±",round(Wild_sd,2)),
           Pre_grass=str_c(round(Pre_grass_mean,2),"±",round(Pre_grass_sd,2)),
           Post_grass=str_c(round(Post_grass_mean,2),"±",round(Post_grass_sd,2))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(family,Total,Wild,Pre_grass,Post_grass) %>% 
    paged_table()
```

```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~diet)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")
```

### Genus

```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus, diet) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__") %>%
  mutate(genus= sub("^g__", "", genus))
genus_summary$diet <- factor(genus_summary$diet, levels=c("Pre_grass", "Post_grass", "Wild"))
```

### origin and diet
```{r taxonomy_genera_summary_oridiet, warning=FALSE, comments="", message=FALSE}
genus_summary %>%
    group_by(genus) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Wild_mean=mean(relabun[diet=="Wild"]*100, na.rm=T),
              Wild_sd=sd(relabun[diet=="Wild"]*100, na.rm=T),
              Pre_grass_mean=mean(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Pre_grass_sd=sd(relabun[diet=="Pre_grass"]*100, na.rm=T),
              Post_grass_mean=mean(relabun[diet=="Post_grass"]*100, na.rm=T),
              Post_grass_sd=sd(relabun[diet=="Post_grass"]*100, na.rm=T))  %>%
    mutate(Total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Wild=str_c(round(Wild_mean,2),"±",round(Wild_sd,2)),
           Pre_grass=str_c(round(Pre_grass_mean,2),"±",round(Pre_grass_sd,2)),
           Post_grass=str_c(round(Post_grass_mean,2),"±",round(Post_grass_sd,2))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(genus,Total,Wild,Pre_grass,Post_grass) %>% 
    paged_table()
```

```{r taxonomy_jitterplot_genus, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary_sort <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) 

genus_summary %>%
  mutate(genus=factor(genus, levels=rev(genus_summary_sort %>% pull(genus)))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
  scale_color_manual(values=phylum_colors) +
  geom_jitter(alpha=0.5) + 
  facet_grid(.~diet)+
  theme_minimal() + 
  theme(axis.text.y = element_text(size=6))+
  labs(y="Family", x="Relative abundance", color="Phylum")

```

