# Community composition

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE, eval=FALSE}
load("data/data.Rdata")

sample_metadata <- read_csv("data/sample_metadata.csv")
 
sample_metadata$individual <- sub("[0-9]{2}$", "", sample_metadata$animal)

sample_metadata_illu <- read_csv("data/metadata_illu.csv")
sample_metadata <- sample_metadata %>% 
  left_join(sample_metadata_illu[c(1,4,9)], by=join_by("sample"=="DogID")) 

load("data/beta.Rdata")
#samples to remove
samples_to_remove <- sample_metadata %>% 
  filter(individual == "GL_02_" | Samples %in% c("Blank", "Saliva") | group2 %in% c("Unknown") | season=="winter" ) %>% #!day %in% c("day_1", "Unknown")
  select(sample) %>% 
  pull()

sample_metadata <- sample_metadata %>% 
  filter(!sample %in% samples_to_remove)
  #filter(individual != "GL_02_" | !Samples %in% c("Blank", "Saliva") | group2 != "Unknown" | season!="winter") 

genome_counts_filt <- genome_counts_filt %>% 
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>% 
  semi_join(., genome_counts_filt, by = "genome") %>% 
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ country, scale="free", space = "free")+
 #   facet_nested(. ~ region+diet,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Bacteria phyla in wild individuals**

```{r phyla_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}

wild_samples <- sample_metadata %>% 
  filter(region=="Gipuzkoa") %>% 
  dplyr::select(sample) %>% 
  pull()

wild_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(wild_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% wild_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla captive animals**

```{r phyla_cap, comment="", echo=FALSE, message=FALSE, warning=FALSE}

captive_samples <- sample_metadata %>% 
  filter(region=="Nafarroa") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla before grass is included in the diet**

```{r phyla_pre, comment="", echo=FALSE, message=FALSE, warning=FALSE}

captive_pre_samples <- sample_metadata %>% 
  filter(diet=="Sled_dog") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_pre_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_pre_samples)) %>%
  as.data.frame() %>% 
  mutate(row_sum = rowSums(.)) %>%
  filter(row_sum != 0) %>%
  select(-row_sum)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_pre_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Bacteria phyla after grass is included in the diet**

```{r phyla_post, comment="", echo=FALSE, message=FALSE, warning=FALSE}
captive_post_samples <- sample_metadata %>% 
  filter(diet=="Post_grass") %>% 
  dplyr::select(sample) %>% 
  pull()

captive_post_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(captive_post_samples)) %>%
  as.data.frame() %>% 
  mutate(row_sum = rowSums(.)) %>%
  filter(row_sum != 0) %>%
  select(-row_sum)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% captive_post_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```

**Number of Archaea phyla**

```{r arch, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Archaea phyla in wild individuals**

```{r arch_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% wild_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Archaea phyla before grass is included in the diet**

```{r arch_pre, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% captive_pre_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull()
```

**Archaea phyla after grass is included in the diet**

```{r arch_post, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(genome %in% captive_post_genomes) %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "s__") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull()

```
```{r nonspe_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "s__") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage=100*count_nospecies/count_total) %>% 
  tt()

```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nmags <- nrow(genome_counts)
nonspecies <- genome_metadata %>%
    filter(species == "s__") %>%
    nrow()
perct <- nonspecies*100/nmags
perct
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "g__") %>%
    nrow()
cat(nongenera)
```


### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,group2) %>%
  summarise(relabun=sum(count))
```

#### Origin: Dogs vs Sled dogs vs Wolves
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
    group_by(phylum) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Dog_mean=mean(relabun[group2=="Dog"]*100, na.rm=T),
              Dog_sd=sd(relabun[group2=="Dog"]*100, na.rm=T),
              Sled_dog_mean=mean(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Sled_dog_sd=sd(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Wolf_mean=mean(relabun[group2=="Wolf"]*100, na.rm=T),
              Wolf_sd=sd(relabun[group2=="Wolf"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,3),"±",round(total_sd,3)),
           Dog=str_c(round(Dog_mean,3),"±",round(Dog_sd,3)),
           Sled_dog=str_c(round(Sled_dog_mean,3),"±",round(Sled_dog_sd,3)),
           Wolf=str_c(round(Wolf_mean,3),"±",round(Wolf_sd,3))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(phylum,total,Dog,Sled_dog,Wolf)
```

## Taxonomy boxplot

### Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family, group2) %>%
  summarise(relabun=sum(count))
family_summary$group2 <- factor(family_summary$group2, levels=c("Sled_dog", "Dog", "Wolf"))
```

#### Origin: Dog vs Captive
```{r taxonomy_family_summary_origin, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Dog_mean=mean(relabun[group2=="Dog"]*100, na.rm=T),
              Dog_sd=sd(relabun[group2=="Dog"]*100, na.rm=T),
              Sled_dog_mean=mean(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Sled_dog_sd=sd(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Wolf_mean=mean(relabun[group2=="Wolf"]*100, na.rm=T),
              Wolf_sd=sd(relabun[group2=="Wolf"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,3),"±",round(total_sd,3)),
           Dog=str_c(round(Dog_mean,3),"±",round(Dog_sd,3)),
           Sled_dog=str_c(round(Sled_dog_mean,3),"±",round(Sled_dog_sd,3)),
           Wolf=str_c(round(Wolf_mean,3),"±",round(Wolf_sd,3))) %>% 
    arrange(-Sled_dog_mean) %>% 
    dplyr::select(family,total,Dog,Sled_dog,Wolf)
```

### Genus

```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus, group2) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__") %>%
  mutate(genus= sub("^g__", "", genus))
genus_summary$group2 <- factor(genus_summary$group2, levels=c("Sled_dog", "Wolf", "Dog"))
```
```{r}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  filter(genome=="ERR1915648_bin_35072")%>% #append genome metadata
  group_by(group2) %>%
  summarise(relabun=sum(count))
```
```{r}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  filter(genus=="Cetobacterium_A")%>% #append genome metadata
  group_by(genome, group2) %>%
  summarise(relabun=sum(count), sd=sd(count))
```

```{r taxonomy_genera_summary_origroup2, warning=FALSE, comments="", message=FALSE}
genus_summary %>%
    group_by(genus) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Dog_mean=mean(relabun[group2=="Dog"]*100, na.rm=T),
              Dog_sd=sd(relabun[group2=="Dog"]*100, na.rm=T),
              Sled_dog_mean=mean(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Sled_dog_sd=sd(relabun[group2=="Sled_dog"]*100, na.rm=T),
              Wolf_mean=mean(relabun[group2=="Wolf"]*100, na.rm=T),
              Wolf_sd=sd(relabun[group2=="Wolf"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,3),"±",round(total_sd,3)),
           Dog=str_c(round(Dog_mean,3),"±",round(Dog_sd,3)),
           Sled_dog=str_c(round(Sled_dog_mean,3),"±",round(Sled_dog_sd,3)),
           Wolf=str_c(round(Wolf_mean,3),"±",round(Wolf_sd,3))) %>% 
    arrange(genus) %>% 
    dplyr::select(genus,total,Dog,Sled_dog,Wolf)
```

```{r taxonomy_jitterplot_genus, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary_sort <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) 

genus_summary %>%
  mutate(genus=factor(genus, levels=rev(genus_summary_sort %>% pull(genus)))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
  scale_color_manual(values=phylum_colors) +
  geom_jitter(alpha=0.5) + 
  facet_grid(.~group2)+
  theme_minimal() + 
  theme(axis.text.y = element_text(size=6))+
  labs(y="Family", x="Relative abundance", color="Phylum")

```

