# Important pathways

```{r load_data_starch, comment="", message=FALSE, warning=FALSE}
load("data/data.Rdata")
load("data/genome_gifts_corrected.Rdata")

genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))

remove_sled <- read_csv("data/sample_metadata.csv") %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  mutate(country = if_else(country == "Shangai domestic dog", "Shangai", country)) %>% 
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_type = case_when(
      breed == "Wolf"        ~ "wolf",
      country == "Greenland" ~ "sled_dog",
      TRUE                   ~ "domestic"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```
```{r load_genome_annotation_starch, warning=FALSE, comments="", message=FALSE}
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz") %>%
  mutate(genome = str_extract(gene, "^[^^]+")) %>%
  mutate(ec = str_c("[EC:",ec,"]")) %>% 
  select(genome, everything())
```
```{r distill_annotations_starch, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=1,annotcol=c(6,7,8,9), verbosity=TRUE)
genome_gifts <- genome_gifts[, !grepl("^S", colnames(genome_gifts))]
```
```{r extra_objects_starch, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(genome_gifts, 
     file = "data/genome_gifts_corrected.Rdata")
```

```{r}
# =========================================================
# STARCH DEGRADATION
# =========================================================

starch_annotations <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    cazy %in% c(
      "GH13",
      "GH31",
      "GH77"
    ) |
      
    kegg %in% c(
      "K01176", # alpha-amylase
      "K01208", # alpha-amylase
      "K01200", # pullulanase
      "K01279", # neopullulanase
      "K01187"  # alpha-glucosidase
    ) |
      
    str_detect(
      ec_clean,
      "3\\.2\\.1\\.1|3\\.2\\.1\\.54|3\\.2\\.1\\.41|3\\.2\\.1\\.135|3\\.2\\.1\\.20"
    )
  )

starch_weight <- starch_annotations %>%
  count(genome, name = "n_starch_genes")

starch_matrix <- genome_counts_filt %>%
  filter(genome %in% starch_weight$genome) %>%
  left_join(starch_weight, by = "genome")

starch_weighted <- starch_matrix %>%
  mutate(
    across(
      -c(genome, n_starch_genes),
      ~ . * n_starch_genes
    )
  )

starch_mat <- starch_weighted %>%
  select(-n_starch_genes) %>%
  column_to_rownames("genome") %>%
  as.matrix()

starch_total <- colSums(starch_mat, na.rm = TRUE)

total_microbiome <- genome_counts_filt %>%
  select(-genome) %>%
  colSums(na.rm = TRUE)

starch_relative <- starch_total / total_microbiome

starch_df <- data.frame(
  sample = names(starch_relative),
  starch_index = as.numeric(starch_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

starch_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_fiber = mean(starch_index, na.rm = TRUE),
    sd_fiber = sd(starch_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(starch_index ~ dog_type, data = starch_df)

pairwise.wilcox.test(
  starch_df$starch_index,
  starch_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(starch_df,
       aes(x = dog_type,
           y = starch_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "Starch degradation index",
    title = "Microbial starch degradation potential"
  )
```
## starch-degrading GH13 enzyme system
TRUE amylase core:
3.2.1.54 → alpha-amylase ✔ (this is the key one)
borderline / related starch enzymes:
3.2.1.133 → glucosidase (downstream starch breakdown)
3.2.1.135 → debranching/pullulanase-related
```{r}
amylase_genes <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    kegg == "K01208",
    str_detect(cazy, "^GH13"),
    str_detect(ec_clean, "3\\.2\\.1\\.1|3\\.2\\.1\\.54")
  )

amylase_matrix <- genome_counts_filt %>%
  filter(genome %in% amylase_genes$genome)

total_abundance <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  as.matrix() %>% 
  colSums(., na.rm = TRUE)

amylase_total <- amylase_matrix %>%
  column_to_rownames("genome") %>%
  as.matrix()%>% 
  colSums(., na.rm = TRUE)

amylase_rel <- amylase_total / total_abundance

sample_amylase <- data.frame(
  sample = names(amylase_rel),
  amylase_rel = as.numeric(amylase_rel)
) %>%
  left_join(sample_metadata, by = "sample")

kruskal.test(amylase_rel ~ dog_type, data = sample_amylase)

pairwise.wilcox.test(
  sample_amylase$amylase_rel,
  sample_amylase$dog_type,
  p.adjust.method = "BH"
)

type_colors <- c('#41B6C0','#E5D388',"#bb6749")

ggplot(sample_amylase, aes(dog_type, amylase_rel, fill = dog_type, color = dog_type)) +
  geom_boxplot(alpha=0.5) +
  geom_jitter(width = 0.2) +
  scale_color_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
  scale_fill_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
  theme_classic()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
# +
#   geom_text_repel(
#     aes(label = sample)
#   )
```

## Bile salt hydrolase
```{r}
amylase_genes <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    kegg == "K01442" |
    str_detect(ec_clean, "3\\.5\\.1\\.24")
  )

amylase_matrix <- genome_counts_filt %>%
  filter(genome %in% amylase_genes$genome)

total_abundance <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  as.matrix() %>% 
  colSums(., na.rm = TRUE)

amylase_total <- amylase_matrix %>%
  column_to_rownames("genome") %>%
  as.matrix()%>% 
  colSums(., na.rm = TRUE)

amylase_rel <- amylase_total / total_abundance

sample_amylase <- data.frame(
  sample = names(amylase_rel),
  amylase_rel = as.numeric(amylase_rel)
) %>%
  left_join(sample_metadata, by = "sample")

kruskal.test(amylase_rel ~ dog_type, data = sample_amylase)

ggplot(sample_amylase, aes(dog_type, amylase_rel, fill = dog_type)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_classic()
```

## Urease (nitrogen recycling enzyme)
```{r}
urease_genes <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    kegg == "K01428" |
    str_detect(ec_clean, "3\\.5\\.5\\.5")
  )

urease_matrix <- genome_counts_filt %>%
  filter(genome %in% urease_genes$genome)

total_abundance <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  as.matrix() %>% 
  colSums(., na.rm = TRUE)

urease_total <- urease_matrix %>%
  column_to_rownames("genome") %>%
  as.matrix()%>% 
  colSums(., na.rm = TRUE)

urease_relative <- urease_total / total_abundance

urease_df <- data.frame(
  sample = names(urease_relative),
  urease_index = as.numeric(urease_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

urease_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_urease = mean(urease_index, na.rm = TRUE),
    sd_urease = sd(urease_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(urease_index ~ dog_type, data = urease_df)

pairwise.wilcox.test(
  urease_df$urease_index,
  urease_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(urease_df,
       aes(x = dog_type,
           y = urease_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "Urease degradation index",
    title = "Microbial urease degradation potential"
  )
```

## Chitinase
```{r}
chitinase_genes <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    cazy == "GH13" |
    kegg == "K01183")

chitinase_matrix <- genome_counts_filt %>%
  filter(genome %in% chitinase_genes$genome)

total_abundance <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  as.matrix() %>% 
  colSums(., na.rm = TRUE)

chitinase_total <- chitinase_matrix %>%
  column_to_rownames("genome") %>%
  as.matrix()%>% 
  colSums(., na.rm = TRUE)

chitinase_relative <- chitinase_total / total_abundance

chitinase_df <- data.frame(
  sample = names(chitinase_relative),
  chitinase_index = as.numeric(chitinase_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

chitinase_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_chitinase = mean(chitinase_index, na.rm = TRUE),
    sd_chitinase = sd(chitinase_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(chitinase_index ~ dog_type, data = chitinase_df)

pairwise.wilcox.test(
  chitinase_df$chitinase_index,
  chitinase_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(chitinase_df,
       aes(x = dog_type,
           y = chitinase_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "chitinase degradation index",
    title = "Microbial chitinase degradation potential"
  )
```

## Fiber / plant polysaccharide degradation

```{r}
# =========================================================
# 1. DEFINE FIBER-DEGRADATION FUNCTIONS
# =========================================================

fiber_annotations <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    cazy %in% c(
      "GH5",   # cellulases
      "GH10",  # xylanases
      "GH11",  # xylanases
      "GH43",  # hemicellulose/arabinose
      "GH53"   # galactan degradation
    ) |
      
    kegg %in% c(
      "K01179", # endoglucanase
      "K01181", # xylanase
      "K01209", # arabinofuranosidase
      "K05349", # beta-glucosidase
      "K01728"  # pectin lyase
    ) |
      
    str_detect(ec_clean,
               "3\\.2\\.1\\.4|3\\.2\\.1\\.8|3\\.2\\.1\\.21|3\\.2\\.1\\.55|4\\.2\\.2\\.10")
  )

# =========================================================
# 2. COUNT FIBER GENES PER GENOME (OPTIONAL WEIGHTING)
# =========================================================

fiber_weight <- fiber_annotations %>%
  count(genome, name = "n_fiber_genes")

# =========================================================
# 3. FILTER ABUNDANCE MATRIX
# genome_counts_filt:
# rows = bacterial genomes
# columns = dog samples
# =========================================================

fiber_matrix <- genome_counts_filt %>%
  filter(genome %in% fiber_weight$genome) %>%
  left_join(fiber_weight, by = "genome")

# =========================================================
# 4. CREATE WEIGHTED MATRIX
# =========================================================

fiber_mat_weighted <- fiber_matrix %>%
  mutate(
    across(
      -c(genome, n_fiber_genes),
      ~ . * n_fiber_genes
    )
  )

# =========================================================
# 5. CONVERT TO MATRIX
# =========================================================

fiber_mat <- fiber_mat_weighted %>%
  select(-n_fiber_genes) %>%
  column_to_rownames("genome") %>%
  as.matrix()

# =========================================================
# 6. SAMPLE-LEVEL FIBER INDEX
# =========================================================

fiber_total <- colSums(fiber_mat, na.rm = TRUE)

# =========================================================
# 7. NORMALIZE BY TOTAL COMMUNITY ABUNDANCE
# =========================================================

total_microbiome <- genome_counts_filt %>%
  select(-genome) %>%
  colSums(na.rm = TRUE)

fiber_relative <- fiber_total / total_microbiome

# =========================================================
# 8. CREATE FINAL TABLE
# =========================================================

fiber_df <- data.frame(
  sample = names(fiber_relative),
  fiber_index = as.numeric(fiber_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

fiber_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_fiber = mean(fiber_index, na.rm = TRUE),
    sd_fiber = sd(fiber_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(fiber_index ~ dog_type, data = fiber_df)

pairwise.wilcox.test(
  fiber_df$fiber_index,
  fiber_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(fiber_df,
       aes(x = dog_type,
           y = fiber_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "Fiber degradation index",
    title = "Microbial plant polysaccharide degradation potential"
  )
```

```{r}
# =========================================================
# PROTEIN DEGRADATION
# =========================================================

protein_annotations <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    
    pfam %in% c(
      "PF00082", # subtilase
      "PF01432", # peptidase M16
      "PF01546", # peptidase M28
      "PF00675"  # cysteine protease
    ) |
      
    kegg %in% c(
      "K01251", # tripeptidase
      "K01358", # aminopeptidase
      "K07727", # protease
      "K08372"
    ) |
      
    str_detect(
      ec_clean,
      "3\\.4\\."
    )
  )

protein_weight <- protein_annotations %>%
  count(genome, name = "n_protein_genes")

protein_matrix <- genome_counts_filt %>%
  filter(genome %in% protein_weight$genome) %>%
  left_join(protein_weight, by = "genome")

protein_weighted <- protein_matrix %>%
  mutate(
    across(
      -c(genome, n_protein_genes),
      ~ . * n_protein_genes
    )
  )

protein_mat <- protein_weighted %>%
  select(-n_protein_genes) %>%
  column_to_rownames("genome") %>%
  as.matrix()

protein_total <- colSums(protein_mat, na.rm = TRUE)

total_microbiome <- genome_counts_filt %>%
  select(-genome) %>%
  colSums(na.rm = TRUE)

protein_relative <- protein_total / total_microbiome

protein_df <- data.frame(
  sample = names(protein_relative),
  protein_index = as.numeric(protein_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

protein_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_fiber = mean(protein_index, na.rm = TRUE),
    sd_fiber = sd(protein_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(protein_index ~ dog_type, data = protein_df)

pairwise.wilcox.test(
  protein_df$protein_index,
  protein_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(protein_df,
       aes(x = dog_type,
           y = protein_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "Protein degradation index",
    title = "Microbial protein degradation potential"
  )
```
Lipid degradation / fat metabolism index
```{r}
# =========================================================
# LIPID DEGRADATION
# =========================================================

lipid_annotations <- genome_annotations %>%
  mutate(ec_clean = coalesce(ec, "")) %>%
  filter(
    
    kegg %in% c(
      "K00249", # acyl-CoA dehydrogenase
      "K01692", # enoyl-CoA hydratase
      "K00626", # acetyl-CoA acetyltransferase
      "K01442"  # bile salt hydrolase
    ) |
      
    pfam %in% c(
      "PF00441", # acyl-CoA dehydrogenase
      "PF00378", # enoyl-CoA hydratase
      "PF02737"  # thiolase
    ) |
      
    str_detect(
      ec_clean,
      "1\\.3\\.8\\.|4\\.2\\.1\\.17|2\\.3\\.1\\.9|3\\.5\\.1\\.24"
    )
  )

lipid_weight <- lipid_annotations %>%
  count(genome, name = "n_lipid_genes")

lipid_matrix <- genome_counts_filt %>%
  filter(genome %in% lipid_weight$genome) %>%
  left_join(lipid_weight, by = "genome")

lipid_weighted <- lipid_matrix %>%
  mutate(
    across(
      -c(genome, n_lipid_genes),
      ~ . * n_lipid_genes
    )
  )

lipid_mat <- lipid_weighted %>%
  select(-n_lipid_genes) %>%
  column_to_rownames("genome") %>%
  as.matrix()

lipid_total <- colSums(lipid_mat, na.rm = TRUE)

total_microbiome <- genome_counts_filt %>%
  select(-genome) %>%
  colSums(na.rm = TRUE)

lipid_relative <- lipid_total / total_microbiome

lipid_df <- data.frame(
  sample = names(lipid_relative),
  lipid_index = as.numeric(lipid_relative)
) %>%
  left_join(sample_metadata, by = "sample")

# =========================================================
# 9. SUMMARY
# =========================================================

lipid_df %>%
  group_by(dog_type) %>%
  summarise(
    mean_fiber = mean(lipid_index, na.rm = TRUE),
    sd_fiber = sd(lipid_index, na.rm = TRUE),
    n = n()
  )

# =========================================================
# 10. STATISTICS
# =========================================================

kruskal.test(lipid_index ~ dog_type, data = lipid_df)

pairwise.wilcox.test(
  lipid_df$lipid_index,
  lipid_df$dog_type,
  p.adjust.method = "BH"
)

# =========================================================
# 11. VISUALIZATION
# =========================================================

ggplot(lipid_df,
       aes(x = dog_type,
           y = lipid_index,
           fill = dog_type)) +
  
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  
  geom_jitter(
    width = 0.15,
    size = 1,
    alpha = 0.8
  ) +
  
  theme_classic(base_size = 14) +
  
  labs(
    x = "Dog group",
    y = "Lipid degradation index",
    title = "Microbial lipid degradation potential"
  )
```



```{r}
amylase <- genome_annotations %>% filter(cazy=="GH13")

genome_counts_filt %>% 
  mutate(across(-genome, ~ . / sum(.))) %>% 
  filter(genome %in% amylase$genome) %>% 
  column_to_rownames(., "genome") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>% 
  left_join(sample_metadata, by="sample") %>% 
  select(sample, dog_type, total) %>% 
  group_by(dog_type) %>% 
  summarise(total_mean= mean(total))

gh13_per_genome <- genome_annotations %>%
  filter(cazy == "GH13") %>%
  count(genome, name = "GH13_gene_count")

genome_annotations %>%
  filter(cazy == "GH13")%>%
  separate_rows(ec, sep = " ") %>%
  filter(grepl("3.2.1", ec)) %>%
  count(ec, genome)


amylase_index <- genome_annotations %>%
  mutate(
    ec_clean = coalesce(ec, ""),
    
    amylase_flag = if_else(
      cazy == "GH13" |
      kegg == "K01208" |
      str_detect(ec_clean, "3\\.2\\.1"),
      1, 0
    )
  ) %>%
  group_by(genome) %>%
  summarise(amylase_score = sum(amylase_flag, na.rm = TRUE))
  
```



```{r amylase_weight, echo=FALSE, eval=FALSE}

amylase_genes <- amylase_matrix %>%
  mutate(n_amylase = n_amylase_genes)

weighted <- colSums(amylase_genes[, -1] * amylase_genes$n_amylase)


# amylase_weight <- amylase_genes %>%
#   count(genome, name = "n_amylase_genes")

# sample_amylase <- amylase_matrix %>%
#   column_to_rownames("genome") %>%
#   as.matrix() %>%
#   colSums(na.rm = TRUE) %>%
#   as.data.frame() %>%
#   rownames_to_column("sample") %>%
#   rename(amylase_total = ".")

# sample_amylase <- sample_amylase %>%
#   left_join(sample_metadata, by = "sample")

sample_amylase <- genome_counts_filt %>%
    mutate(across(-genome, ~ . / sum(.))) %>% 
  filter(genome %in% amylase_weight$genome) %>% 
  left_join(amylase_weight, by = "genome") %>%
  mutate(n_amylase_genes = replace_na(n_amylase_genes, 0)) %>%
  mutate(across(-c(genome, n_amylase_genes), ~ . * n_amylase_genes))

sample_amylase <- sample_amylase %>%
  t() %>% 
  
  left_join(sample_metadata, by = "sample")

amylase_index <- genome_annotations %>%
  mutate(
    ec_clean = coalesce(ec, ""),
    
    amylase_flag = if_else(
      cazy == "GH13" |
      kegg == "K01208" |
      str_detect(ec_clean, "3\\.2\\.1\\.54"),
      1, 0
    )
  ) %>%
  group_by(genome) %>%
  summarise(amylase_score = sum(amylase_flag, na.rm = TRUE)) %>% 
  filter(amylase_score >0)

amylase_index <- genome_annotations %>%
  mutate(
    ec_clean = coalesce(ec, ""),
    
    amylase_flag = if_else(
      cazy == "GH13" |
      kegg == "K01208" |
      str_detect(ec_clean, "3\\.2\\.1"),
      1, 0
    )
  ) %>%
  group_by(genome) %>%
  summarise(amylase_score = sum(amylase_flag, na.rm = TRUE)) %>% 
  filter(amylase_score >0)

genome_counts_filt %>% 
  mutate(across(-genome, ~ . / sum(.))) %>% 
  filter(genome %in% amylase_index$genome) %>% 
  column_to_rownames(., "genome") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>% 
  left_join(sample_metadata, by="sample") %>% 
  select(sample, dog_type, total) %>% 
  group_by(dog_type) %>% 
  summarise(total_mean= mean(total))

```