```{r load_data_longi, comment="", message=FALSE, warning=FALSE}
load("data/data.Rdata")
load("data/beta.Rdata")
genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))

day_info <- read_csv("data/sample_metadata.csv")

sample_metadata <- sample_metadata %>% 
  left_join(., day_info %>% select(sample, animal, day, season), by="sample") %>% 
  filter(locality == "Ittoqqortoormiit" & season=="winter")%>%
  filter(!day %in% c("day_6", "day_7")) %>% 
  mutate(animal_id = str_remove(animal, "\\d$"))

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```

## Taxonomy overview 

```{r taxonomy_barplot_longi, fig.height=8, fig.width=20, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=animal_id, y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values=phylum_colors) +
  facet_grid(~ day, scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_text(size = 10, angle=45, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black")
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

### Phylum relative abundances
```{r taxonomy_phylum_longi, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum, day) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_longi, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(day, phylum) %>%
  summarise(
    n = sum(!is.na(relabun)),
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = ifelse(n > 1, sd(relabun * 100, na.rm = TRUE), 0),
    .groups = "drop"
  ) %>%
    arrange(desc(total_mean)) %>%   # ← reorder rows only
  mutate(
    total = stringr::str_c(round(total_mean, 3), "±", round(total_sd, 3))
  ) %>%
  dplyr::select(phylum, day, total) %>%
  pivot_wider(names_from = "day", values_from = "total", names_sort = FALSE) %>%
select(phylum, day_1, day_2, day_3, day_4, day_5) %>% 
  tt()
```


## Alpha diversity
```{r alpha_div_longi, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample))
```

```{r alpha_plots_longi, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic", "functional"))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = value, x = day, group=day, color=day, fill=day)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic")), scales = "free", ncol = 5) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size =12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```
```{r plots_paired_longi, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = "sample") %>%
  ggplot(aes(x = day, y = value, group = animal_id, color = day)) +
  geom_boxplot(aes(group = day, fill = day), alpha = 0.5, outlier.shape = NA) +
  geom_line(alpha = 0.4) +   # ← connects same animal across days
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(
    ~ factor(metric, labels = c(
      "richness" = "Richness",
      "neutral" = "Neutral",
      "phylogenetic" = "Phylogenetic")),
    scales = "free",
    ncol = 5
  ) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

```{r alpha_div_diets_summary_longi, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(day, alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T)
  ) %>%
  mutate(
    total = str_c(round(total_mean, 3), "±", round(total_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(alpha, day, total) %>%
  pivot_wider(names_from = alpha, values_from = total) %>% 
  tt()
```


## Beta diversity
```{r}
#Neutral
samples_to_keep <- sample_metadata %>%
  select(sample) %>% 
  pull()
neutral <- as.matrix(beta_q1n$S)
neutral_sub <- neutral[samples_to_keep, samples_to_keep]
neutral_dis <- as.dist(neutral_sub)

# sample_metadata_ordered <- sample_metadata %>% 
#   arrange(match(sample,labels(neutral_dis))) %>% 
#   column_to_rownames(., "sample")

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% samples_to_keep) %>%
  column_to_rownames("sample") %>%
  .[labels(neutral_dis), ]

sample_metadata_ordered_row <- sample_metadata %>% 
  arrange(match(sample,labels(neutral_dis)))

disp <- betadisper(neutral_dis, sample_metadata_ordered$day)
anova(disp)

adonis2(
  neutral_dis ~ day,
  data = sample_metadata_ordered,
  permutations = how(blocks = sample_metadata_ordered$animal_id)
)
```

```{r neutral_animal_id_plot, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
neutral_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(animal_id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = animal_id)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  )

# +
#   geom_text_repel(aes(label = animal_id), size = 3)

```
```{r neutral_day_plot_longi, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
neutral_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(day) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = day)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  )
```
```{r}
cca_ord <- capscale(neutral_dis ~ day, data = sample_metadata_ordered)

CAP_df <- as.data.frame(vegan::scores(cca_ord, display = "sites")) %>%
  rownames_to_column('sample') %>%
  left_join(sample_metadata_ordered_row, by = 'sample') %>%
  column_to_rownames('sample')%>%
  mutate(x_cen = mean(CAP1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(CAP2, na.rm = TRUE))

CAP_df %>%
  ggplot(aes(CAP1, CAP2, color = day, group = animal_id)) +
  geom_point() +
  geom_path(alpha = 0.4) +
  theme_classic()
```

# temporal divergence
Microbiome divergence from baseline increased significantly over time (day effect: p < 0.001)
```{r}
meta <- sample_metadata_ordered %>%
  rownames_to_column("sample")

baseline_df <- meta %>%
  filter(day == "day_1") %>%
  select(animal_id, baseline_sample = sample)

df_dist <- meta %>%
  left_join(baseline_df, by = "animal_id") %>%
  mutate(dist = as.matrix(neutral_dis)[cbind(sample, baseline_sample)]) %>%
  mutate(
    day_num = as.numeric(gsub("day_", "", day)),
    animal_id = factor(animal_id)
  )

m1 <- lmerTest::lmer(dist ~ day_num + (1 | animal_id), data = df_dist) #Do individuals diverge from their baseline over time? Does distance from baseline (day 1) change over time, while accounting for repeated measures within each animal?
summary(m1)
anova(m1)
#day_num = 0.0866, p = 1.76e-05. On average, microbiome distance from baseline increases significantly over time
```

# do animals respond differently over time?
Inclusion of random slopes did not improve model fit and resulted in singularity, indicating consistent temporal trajectories across individuals.
```{r}
m2 <- lmerTest::lmer(dist ~ day_num + (day_num | animal_id), data = df_dist)
isSingular(m2)
summary(m2)
anova(m1)
```

```{r}
anova(m1, m2)
```

Divergence occurred early (day 1–2) and then plateaued
```{r}
m1_cat <- lmer(dist ~ day + (1 | animal_id), data = df_dist)
emmeans(m1_cat, pairwise ~ day)
```

```{r}
ggplot(df_dist, aes(day_num, dist, group = animal_id)) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(group = 1), fun.data = mean_se, geom = "errorbar", width = 0.1) +
  theme_classic()
```

